**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-06-12 submission/manuscript.Rmd
more references

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation prior to the procedure (Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research. ((Is anything statistically significant here? use Doherty's code)). Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on. Thus there was no clear pattern of changes in microbial diversity along the gut axis.

To determine if the structure of the community is consistent between locations we constructed an NMDS ordination to examine the distances between each sample. We found that microbial communities do not cluster on an ordination by the left or right sides in either the mucosal or lumenal samples (Figure 2C). Previous human microbiome studies that have sampled multiple sites have found that individuals are typically more similar to themselves than others in the study (cite). To see if this effect was driving the organization of the ordination, we connected communities from the same individual using gray lines on Figure 2C. While some individuals did have samples that clustered near each other (shorter lines), there were just as many individuals who's samples diverged from each other (longer lines). Similar to the diversity metrics, when examining community structure between sites and sides, there is no clear pattern between or within individuals.

To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals. Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment). Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool. Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Further, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). This finding suggests that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different. ((mention the other 2 comparisons? "among the four comparisons made..."))

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt? need to follow up with this too)) features for the left and ((Xopt)) for the right. The models performed very well when classifying these samples (AUC X and Y). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

Next, we built a model to differentiate the left and right microbiome. The model performed best when distinguishing the left versus right mucosa (Figure 5A) compared to the left versus right lumen. The model was able to explain ((X%)) of the variance. OTUs that were differentially abundant between the left and right mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). The model was less-effective at classifying the right and left lumen ((AUC)). However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also drive sidedness- such as _Anaerococcus_ and _Finegoldia_.

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Some of the healthy patients have Fusobacterium OTUs. The most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_. It is the only OTU to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. ((cite joe and niels work somewhere in here, maybe in the discussion?)). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample. ((Table or something?)). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides. Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248. Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa. Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.

#### Exit stool comparisons model - specifics here or above?

>>>>

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation prior to the procedure (Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research. Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on. Thus there was no clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals. Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment). Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool. Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Further, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005). Thus interpersonal variation between subjects drives the differences between samples more than sample site or location. Overall, the results of the similarity analysis suggest that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left. The models performed very well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

Next, we built a model to differentiate the left and right microbiome. The model performed best when distinguishing the right versus left mucosa (Figure 5A, AUC = 0.912) compared to the right versus left lumen (AUC = 0.755). The model was able to explain ((X%)) of the variance. OTUs that were differentially abundant between the left and right mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). The model was less-effective at classifying the right and left lumen (AUC = 0.755). However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also drive sidedness- such as _Anaerococcus_ and _Finegoldia_.

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_. It was the only sequence to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample (Figure 6A). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides (Figure 6B). Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248. Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa (Figure 6C). Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.
